---
title: "flexhaz: Hazard-First Survival Modeling"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{flexhaz: Hazard-First Survival Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## The Problem

Standard parametric survival models force you to choose from a short list of
hazard shapes: constant (exponential), monotone power-law (Weibull), exponential
growth (Gompertz). But real failure patterns rarely cooperate:

- **Ceramic capacitors** exhibit wear-out that accelerates with age — the failure
  rate climbs faster than any power law.
- **Software systems** follow a bathtub curve: early crashes from latent bugs,
  a stable period, then increasing failures as dependencies rot.
- **Post-surgical patients** face high initial risk that drops sharply, then
  slowly rises again as long-term complications emerge.

Each of these demands a hazard function that standard families cannot express.
You can reach for semi-parametric methods (Cox regression), but then you lose the
ability to extrapolate, simulate, or compute closed-form quantities like mean
time to failure.

## The Idea

`flexhaz` takes a different approach: **you write the hazard function, and the
package derives everything else**.

Given a hazard (failure rate) function $h(t)$, the full distribution follows:

$$
h(t) \;\xrightarrow{\;\int\;}\; H(t) = \int_0^t h(u)\,du
      \;\xrightarrow{\;\exp\;}\; S(t) = e^{-H(t)}
      \;\xrightarrow{\;\times h\;}\; f(t) = h(t)\,S(t)
$$

If you can express $h(t)$ as an R function of time and parameters, `flexhaz`
gives you survival curves, CDFs, densities, quantiles, random sampling,
log-likelihoods, MLE fitting, and residual diagnostics — automatically.

## A Complete Example

Let's model a system with a **bathtub-shaped** failure rate: high early risk
(infant mortality), a stable useful-life period, and accelerating wear-out.

```{r setup}
library(flexhaz)
```

### Define the hazard

$$h(t) = a\,e^{-bt} + c + d\,t^k$$

Three additive components: exponential decay (infant mortality), constant
baseline (useful life), and power-law growth (wear-out).

```{r define-dist}
bathtub <- dfr_dist(
  rate = function(t, par, ...) {
    a <- par[[1]]  # infant mortality magnitude
    b <- par[[2]]  # infant mortality decay rate
    c <- par[[3]]  # baseline hazard
    d <- par[[4]]  # wear-out coefficient
    k <- par[[5]]  # wear-out exponent
    a * exp(-b * t) + c + d * t^k
  },
  par = c(a = 0.5, b = 0.5, c = 0.01, d = 5e-5, k = 2.5)
)
```

### Plot the hazard curve

```{r plot-hazard, fig.alt="Bathtub hazard curve showing high infant mortality at left, flat useful life in the middle, and rising wear-out on the right."}
h <- hazard(bathtub)
t_seq <- seq(0.01, 30, length.out = 300)
plot(t_seq, sapply(t_seq, h), type = "l", lwd = 2, col = "darkred",
     xlab = "Time", ylab = "Hazard rate h(t)",
     main = "Bathtub Hazard Function")
```

### Plot the survival curve

```{r plot-surv, fig.alt="Survival curve derived from the bathtub hazard, showing rapid early decline that stabilizes then drops again."}
plot(bathtub, what = "survival", xlim = c(0, 30),
     col = "darkblue", lwd = 2,
     main = "Survival Function S(t)")
```

### Simulate data and fit via MLE

```{r simulate-fit, warning=FALSE}
# Generate failure times with right-censoring at t = 25
set.seed(42)
samp <- sampler(bathtub)
raw_times <- samp(80)
censor_time <- 25

df <- data.frame(
  t = pmin(raw_times, censor_time),
  delta = as.integer(raw_times <= censor_time)
)

cat("Observed failures:", sum(df$delta), "out of", nrow(df), "\n")
cat("Right-censored:", sum(1 - df$delta), "\n")

# Fit the model
solver <- fit(bathtub)
result <- solver(df,
  par = c(0.3, 0.3, 0.02, 1e-4, 2),
  method = "Nelder-Mead"
)
```

### Inspect the fit

```{r inspect}
cat("Estimated parameters:\n")
print(coef(result))
```

### Residual diagnostics

```{r qq-plot, fig.alt="Cox-Snell residuals Q-Q plot for the bathtub model."}
fitted_bathtub <- dfr_dist(
  rate = function(t, par, ...) {
    par[[1]] * exp(-par[[2]] * t) + par[[3]] + par[[4]] * t^par[[5]]
  },
  par = coef(result)
)
qqplot_residuals(fitted_bathtub, df)
```

Points near the diagonal indicate the model captures the failure pattern well.

## Built-in Distributions

For common parametric families, the package provides optimized constructors
with analytical cumulative hazard and score functions:

| Constructor | Hazard Shape | Parameters |
|-------------|-------------|------------|
| `dfr_exponential(lambda)` | Constant | failure rate |
| `dfr_weibull(shape, scale)` | Power-law | shape $k$, scale $\sigma$ |
| `dfr_gompertz(a, b)` | Exponential growth | initial rate, growth rate |
| `dfr_loglogistic(alpha, beta)` | Non-monotonic (hump) | scale, shape |

Use these when a standard family fits your problem — they include analytical
derivatives for faster, more accurate MLE.

## Who Is This For?

**Reliability engineers**: If you think in failure rates, MTBF, and bathtub
curves, `flexhaz` speaks your language. Define the hazard function that matches
your failure physics, fit it to test data with censoring, and extract B-life
estimates, warranty predictions, and maintenance intervals — all from a single
object.

**Statisticians and survival analysts**: If you work with censored lifetime data
and need more flexibility than Weibull or Cox regression, `flexhaz` gives you
fully parametric models with arbitrary hazard shapes. You get proper likelihood
inference (score, Hessian, confidence intervals) and residual diagnostics, with
the option to supply analytical derivatives or fall back to numerical methods.

## Where to Go Next

1. **`vignette("getting_started")`** — 5-minute quick start with fitting and diagnostics
2. **`vignette("reliability_engineering")`** — Five real-world case studies
3. **`vignette("failure_rate")`** — Mathematical foundations of hazard-based modeling
4. **`vignette("custom_distributions")`** — The three-level optimization paradigm
5. **`vignette("custom_derivatives")`** — Analytical score and Hessian functions
