---
title: "Getting Started with dfr.dist"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with dfr.dist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 5-Minute Quick Start

**The Scenario**: You have failure-time data from 50 devices. Some failed during
testing; others were still running when the test ended (right-censored). You
suspect the failure rate increases over time. You need to: (1) fit a model,
(2) estimate parameters with uncertainty, (3) verify the fit.

The `dfr.dist` package lets you work with survival distributions through their
**hazard functions** (also called failure rate functions). This gives you complete
flexibility to model systems with time-varying failure patterns.

### Installation

```{r eval=FALSE}
# From GitHub
remotes::install_github("queelius/dfr.dist")
```

### Load the Package

```{r setup}
library(dfr.dist)
```

## Your First DFR Distribution

Let's start with the simplest case: the exponential distribution with constant
failure rate.

```{r}
# Create an exponential distribution with failure rate lambda = 0.1
# (Mean time to failure = 10 time units)
exp_dist <- dfr_exponential(lambda = 0.1)

# Print summary
print(exp_dist)
```

### Get Distribution Functions

All distribution functions are accessed through a two-step pattern:

```{r}
# 1. Get the function
S <- surv(exp_dist)        # Survival function
h <- hazard(exp_dist)      # Hazard (failure rate)
f <- density(exp_dist)     # PDF

# 2. Evaluate at specific times
S(10)   # P(survive past t=10) = exp(-0.1 * 10) ≈ 0.37
h(10)   # Hazard at t=10 = 0.1 (constant for exponential)
f(10)   # PDF at t=10
```

### Generate Samples

```{r}
# Create a sampler and generate failure times
samp <- sampler(exp_dist)
set.seed(42)
failure_times <- samp(20)

# Check: sample mean should be close to MTBF = 1/lambda = 10
mean(failure_times)
```

## Fitting to Data

Now let's fit a model to survival data.

```{r}
# Simulate failure data (in practice, you'd have real data)
set.seed(123)
n <- 50
df <- data.frame(
  t = rexp(n, rate = 0.08),   # True lambda = 0.08
  delta = rep(1, n)           # All exact observations
)

head(df)
```

### Maximum Likelihood Estimation

```{r}
# Create distribution template (no parameters)
exp_template <- dfr_exponential()

# Create solver and fit
solver <- fit(exp_template)
result <- solver(df, par = c(0.1))  # Initial guess

# Extract fitted parameter
coef(result)

# Compare to analytical MLE: lambda_hat = n / sum(t)
n / sum(df$t)
```

### Model Diagnostics

Check if the model fits well using residuals:

```{r, fig.alt="Cox-Snell residuals Q-Q plot showing points close to diagonal line, indicating good fit."}
# Create fitted distribution
fitted_dist <- dfr_exponential(lambda = coef(result))

# Q-Q plot of Cox-Snell residuals (should follow Exp(1) line)
qqplot_residuals(fitted_dist, df)
```

## Working with Censored Data

Real survival data often includes **censoring** - observations where we only know
the subject survived past a certain time.

```{r}
# Data with right-censoring (delta=0 means censored)
df_censored <- data.frame(
  t = c(5, 8, 12, 15, 20, 25, 30, 30, 30, 30),
  delta = c(1, 1, 1, 1, 1, 0, 0, 0, 0, 0)  # Last 5 censored at t=30
)

# Fit handles censoring automatically
solver <- fit(dfr_exponential())
result <- solver(df_censored, par = c(0.1))
coef(result)
```

## Beyond Exponential: Weibull Distribution

The exponential assumes constant failure rate. For **increasing** or **decreasing**
failure rates, use Weibull:

```{r, fig.alt="Two hazard curves: exponential (flat line) and Weibull shape=2 (increasing curve)."}
# Weibull with increasing failure rate (wear-out)
weib_dist <- dfr_weibull(shape = 2, scale = 20)

# Compare hazard functions
plot(weib_dist, what = "hazard", xlim = c(0, 30), col = "blue",
     main = "Hazard Comparison")
plot(dfr_exponential(lambda = 0.05), what = "hazard", add = TRUE, col = "red")
legend("topleft", c("Weibull (shape=2)", "Exponential"),
       col = c("blue", "red"), lwd = 2)
```

Weibull shape parameter interpretation:

- **shape < 1**: Decreasing hazard (infant mortality, burn-in failures)
- **shape = 1**: Constant hazard (exponential)
- **shape > 1**: Increasing hazard (wear-out, aging)

## Next Steps

Now that you understand the basics, explore:

1. **`vignette("reliability_engineering")`** - Five real-world case studies
2. **`vignette("failure_rate")`** - Mathematical foundations
3. **`vignette("custom_distributions")`** - Build optimized custom distributions
4. **`vignette("automatic_differentiation")`** - Analytical derivatives for MLE

### Available Distributions

| Constructor | Hazard Pattern | Use Case |
|-------------|----------------|----------|
| `dfr_exponential()` | Constant | Random failures, useful life |
| `dfr_weibull()` | Power-law (↑ or ↓) | Wear-out, infant mortality |
| `dfr_gompertz()` | Exponential growth | Biological aging |
| `dfr_loglogistic()` | Non-monotonic (↑ then ↓) | Initial risk that diminishes |

### Key Functions

| Function | Purpose |
|----------|---------|
| `hazard()` | Get hazard (failure rate) function |
| `surv()` | Get survival function S(t) |
| `density()` | Get PDF f(t) |
| `fit()` | Maximum likelihood estimation |
| `residuals()` | Model diagnostics |
| `plot()` | Visualize distribution functions |
