---
title: "Exponential distribution"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exponential distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

We load the R package `algebraic.dist` which contains all of the
relevant material in this section with:

```{r setup, message = FALSE}
if (!require(algebraic.dist)) {
  devtools::install_github("queelius/algebraic.dist")
}
library(algebraic.dist)
```

## Exponential distribution

A random variable $T$ is said to have an Exponential
distribution with parameter $\lambda$ if its probability density
function is given by $f(t|\lambda) = \lambda e^{-\lambda t} 1_{t \geq 0}$,
where $1_{\text{predicate}} = 1$ if the *predicate* condition is true,
otherwise it maps to $0$. The cdf and survival function are respectively
given by $F(t|\lambda) = 1_{t > 0} (1 - e^{-\lambda t})$ and
$S(t|\lambda) = 1 - F(t|\lambda) = 1_{t \leq 0} + 1_{t > 0} e^{-\lambda t}$.

It's hazard function $h(t)$ is given by
$$
h(t|\lambda) = \frac{f(t|\lambda)}{S(t|\lambda)} = 1_{t > 0} \, \lambda,
$$
i.e., it has a constant failure rate $\lambda$ over its entire lifetime.
This is known as memoryless property,
$$
  \Pr\{T > t + \Delta t | T > t\} = \Pr\{T > \Delta t\}
$$
for all $t, \Delta t > 0$.

This is often a reasonable assumption for many systems, such as
electronic components, where the failure rate is essentially constant for
a significant portion of the component's lifetime.

We define the parameters of the i.i.d. random sample with:
```{r init_params}
n <- 1000
rate <- .5
```

We generate a random sample $X_i \sim \operatorname{EXP}(\lambda=`r rate`)$ for
$i=1,\ldots,n$ with:
```{r load_dist, message=F, warning=F}
exp.dist <- exp_dist(rate = rate)
data <- sampler(exp.dist)(n)
```

We have observed a sample of size $n=`r n`$.
We show some observations from this sample with:
```{r print_data}
head(data)
```

We show a histogram of the sample overlaid with a plot of the exponential
function's pdf with:
```{r plotter, fig.width=6, fig.height=4, fig.align="center"}
curve(pdf(exp.dist)(x), col = "red", xlim=c(0 , 10))
hist(data, add = TRUE, freq = FALSE, breaks = 40)
```

We construct a data frame showing mean, variance, and $E(sin(T))$,
the expected value of the sin of $T$. We show how to do it using
the `expectation` API, and then show the sample stimates.

```{r}
df <- data.frame(
  "true value" = c(mean(exp.dist),
                   vcov(exp.dist),
                   (expectation(exp.dist, sin)$value)),
  "sample estimate" = c(mean(data), var(data), mean(sin(data))))
row.names(df) <- c("mean", "variance", "E(sin(T))")
print(df)
```

We provide a general API that allows a number of different operations on
`dist` objects like `exp_dist`, e.g., to obtain the infimum, write:
```{r}
infimum(sup(exp.dist))
```

See the documentation for more details.
